---
title: |
  <center> [EM1415] - Predictive Business and Finance </center>
subtitle: | 
  <center> The Tourism Forecasting 2010 Competition (JIF) </center>
author: |
  <div style="text-align: left; margin: 0 auto; width: fit-content;">
    Crapanzano Edoardo – 889685
  </div>
date: | 
  <center>`r Sys.Date()`</center>
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: show
    self_contained: true
    fig_caption: true
    highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# [Introduction]{style="font-size:20px;"}

Tourism is one of the most rapidly growing global industries and tourism
forecasting is becoming an increasingly important activity in planning
and managing the industry. The 2010 tourism forecasting competition had
two parts. In part I, competitors were tasked with producing forecasts
for the next four years, given 518 series of annual tourism data. The
data consists of 518 annual series, each related to some (undisclosed)
tourism activity. Tourism activities include inbound tourism numbers to
one country from another country, visitor nights in a particular
country, tourism expenditure, etc. The series differ in length, ranging
from 7-year series to 43-year series. They also differ in the order of
magnitude of values. The data is available at
www.kaggle.com/c/tourism1/Data. While the competition goal was to
achieve the highest predictive accuracy, the goal of this case is to
highlight different aspects of the forecasting process that pertain to
forecasting a large number of series.

## [Uploading Data]{style="font-size:20px;"}

```{r tourism data, results = "hide"}
tourism_data = read.csv("tourism_data.csv")
```

The packages used to create the codes are the following:

```{r libraries, results = "hide"}
if (!require("tidyverse")) install.packages("tidyverse") # Collection of data science packages
if (!require("fpp3")) install.packages("fpp3")
if (!require("visdat")) install.packages("visdat")

library(tidyverse)
library(fpp3)
library(visdat)
library(readr)
library(dplyr)          # Data manipulation with verbs
library(ggplot2)        # Create elegant data visualizations
library(tidyr)          # Tidy and reshape data
```

# [Exercise 1]{style="font-size:20px;"}

We need to create the 'Tsibble' object (Annual Time Series) and
transform the data because they are in wide format (columns from Y1 to
Y518). Thus, we need long format for fpp3. We also preserve the time
alignment by adding a Time index before pivoting.

```{r tsibble}
tourism_ts = tourism_data %>%
  mutate(Time = row_number()) %>%
  pivot_longer(cols = starts_with("Y"), names_to = "Series", values_to = "Value") %>%
  filter(!is.na(Value)) %>% # Removing NA values so that the data is clean and ready for the analysis
  as_tsibble(index = Time, key = Series) # Convert to a tsibble (time series tibble)

glimpse(tourism_ts)
```

<br>

Based on the dataset of `tourism_data`, we can investigate the
distribution of all the different Time Series or a random subset of
them.

**Advanced Visualization**: plotting 518 series on one legend is messy.
So we can create a "Spaghetti Plot" by hiding the legend and using
transparency, guaranteeing a clearer view.

```{r plot all ts}
tourism_ts %>%
  autoplot(Value, alpha = 0.6) +
  theme(legend.position = "none") +
  labs(title = "Overview of all Tourism Series", subtitle = "518 series with varying lengths and trends", y = "Tourism metric", x = "Time Index")

tourism_ts %>%
  autoplot(sqrt(Value), alpha = 0.6) +
  theme(legend.position = "none") +
  labs(title = "Overview of all Tourism Series", subtitle = "518 series scaled by the square root", y = "Square root tourism metric", x = "Time Index")

tourism_ts %>%
  model(STL(Value)) %>%
  components()%>%
  autoplot() +
  theme(legend.position = "none") +
  labs(title = "Decomposition of all the Series", subtitle = "Top: data, Middle: trend, Bottom: remainder(noise+cycle)", y = "Tourism metric")

tourism_ts %>%
  model(STL(sqrt(Value))) %>%
  components()%>%
  autoplot() +
  theme(legend.position = "none") +
  labs(title = "Decomposition of all the Series", subtitle = "Top: square root data, Middle: trend, Bottom: remainder(noise+cycle)", y = "Square root tourism metric")
```

<br> The series have different starting points and they differ
significantly in length and in magnitude (tourism activities with
different scales). Type of components that are visible:<br> - *Trend*:
this is the most dominant component across the series and the plot
reveals that most series exhibit strong trends over time. Indeed, the
tourism numbers follow economic expansion, leading to long-term upward
(or occasionally downward) slopes;<br> - *Remainder*: random
fluctuations are present but do not obscure the underlying trends in the
majority of the larger series.<br>

The *cyclical component* is not clearly visible since the remaining
residuals resemble irregular noise rather than a smooth, wave-like
cyclical pattern. With many series being very short (e.g., 7 to 15
years), it is extremely difficult to statistically separate a multi-year
cycle from the trend or random noise.<br> Clear *seasonal patterns* are
difficult to distinguish from the trend for every series. The data
represents annual figures, where seasonality is not applicable because
it is impossible to see intra-year patterns such us quarterly, monthly
or weekly patterns.<br> <br><br>

**Detailed Visualization**: Facet Plot of a random subset of series to
see components clearly, we plot a random sample of 12 series.

```{r facets}
set.seed(987)
sample_series = sample(unique(tourism_ts$Series), 12)

tourism_ts %>%
  filter(Series %in% sample_series) %>%
  autoplot(Value) +
  facet_wrap(~Series, scales = "free_y") +
  theme(legend.position = "none") +
  labs(title = "Detailed view of 12 random Series", subtitle = "Notice the strong trends and differing scales", y = "Tourism metric")
```

<br>

To investigate if the cyclical component is visible or not, we will use
LOESS smoother (STL decomposition) which separates the Trend,
Seasonality, and Remainder (Noise/Cycle). Since the data is annual
(frequency = 1), standard seasonal decomposition doesn't apply
automatically, so we use a local regression (LOESS) to separate the
Trend from the Remainder. If the "Remainder" looks random, there is no
cycle. If it looks like a wave, there is a cycle.

```{r cyclical component}
dcmp = tourism_ts %>%
  filter(Series %in% sample_series)%>%
  model(STL(Value ~ trend(window = 5), robust = TRUE)) %>% # trend controls wiggliness of trend component
  components()

dcmp %>%
  autoplot() +
  labs(title = "Decomposition of randomly selected Series", subtitle = "Top: data, Middle: trend, Bottom: remainder(noise+cycle)", y = "Tourism value")
```

<br> Since most of the series are short, decomposition algorithms tend
to absorb any long-term movement into the trend component. We have
applied a small window (e.g., window = 5) so that the model only looks
at 5 time steps in order to react quickly to changes and bend to follow
the data more closely. When we mathematically remove the trend
(detrending), the remaining "residuals" mostly resemble irregular noise
rather than a smooth, wave-like cyclical pattern. Most series exhibit a
strong, monotonic upward: this overpowering movement makes it difficult
to see the "business cycle" (the rise and fall of tourism demand due to
economic conditions for non-fixed period of at least 2 years). Thus, the
cyclical component is not clearly visible. <br><br>

Then, we need to check for internal missing values through `scan_gaps()`
that searches if there are missing time steps within the series range.

```{r check NA and errors}
#Missing values
gaps = scan_gaps(tourism_ts)

if(nrow(gaps) > 0) {
  print("Warning: Internal missing values found!")
  print(gaps)
} else {
  print("Check passed: No internal missing values found.")
}

#Zero values
total_zeros = sum(tourism_data == 0, na.rm = TRUE)

if (total_zeros > 0) {
  print(paste("ALERT: Found", total_zeros, "zero values."))
} else {
  print("Clean: No zero values found in the dataset.")
}

#Negative values
total_negatives = sum(tourism_data < 0, na.rm = TRUE)

if (total_negatives > 0) {
  print(paste("ALERT: Found", total_negatives, "negative values."))
} else {
  print("Clean: No negative values found.")
}
```

There are no internal missing values (gaps in the middle of a series).
The NA values observed in the dataset are solely due to the differing
lengths and starting points of the series. There are no zero values and
no negative values, which is consistent with tourism flow data. The data
appears clean, though the short duration of some series (e.g., 7
points/years) limits the complexity of models that can be fitted to
them. <br>

```{r plot NA}
vis_miss(tourism_data) +
  labs(title = "Time Series Structure Map",
       subtitle = "Black areas indicate missing values/not yet started series (varying lengths)",
       x = 'Time Series', y = "Years (Time Index)")+
  theme_minimal() +
  theme(axis.text.x = element_blank(), panel.grid = element_blank(), legend.position = "bottom")
```

<br> This plot clearly shows the "stepped" structure of the dataset:
black areas indicate years where the series had not started yet. It does
not mean that tourism was zero in those years; it simply means the data
was not recorded. <br>

# [Exercise 2]{style="font-size:20px;"}

```{r sets}
# Create Training Set (Remove the last 4 years)
n_years = nrow(tourism_data)
train_set = tourism_data[1:(n_years - 4), ]

# Create Validation Set (Keep only the last 4 years)
validation_set = tourism_data[(n_years - 3):n_years, ]

# Verify the split
cat("Training Years:", nrow(train_set), "\n") # Should be 39
cat("Validation Years:", nrow(validation_set), "\n") # Should be 4

#alternative solution using the cleaned dataset in the long format
train_tourism_ts = tourism_ts%>% #training set
  filter(Time <= max(Time) - 4)

test_tourism_ts = tourism_ts%>% #test set
  filter(Time > max(Time) - 4)
```

The logic is *Out-of-Sample Evaluation*: when forecasting, we are always
standing in the present trying to predict the unknown future. By hiding
the last 4 years from the model during training, we simulate this
reality by asking the model: "If you were standing 4 years ago, could
you have accurately predicted what actually happened?"<br> For yearly
data, a 4-year horizon represents a medium-term forecast. This partition
specifically tests if our model captures the underlying trend stability
rather than just short-term noise.

The main disadvantage is Information Loss, which is particularly acute
with yearly data:<br> -the most valuable data for forecasting next year
is usually this year's data. By cutting off the last 4 years, we are
forcing the model to forecast based on data that is already 4 years old.
If a major structural change (e.g., a new travel policy or economic
shift) occurred in the last 3 years, the model will be completely
unaware of it and likely produce poor forecasts.<br> -In our dataset
`tourism_data` we have time series of different length; removing 4
observations is disproportionately harmful to shorter series. For a
series with only 7 observations, this approach leaves only 3 data points
for training (a 57% data loss), rendering statistical estimation of
trends or parameters extremely unstable. In time series analysis, every
data point counts; reducing the training data to a range of 39-3
observations might make it harder for the model to statistically
distinguish between a real trend and random variation. This leaves the
model with very little "training" data to learn from, potentially
leading to underfitting. <br>

# [Exercise 3]{style="font-size:20px;"}

```{r naive forecasts}
# Last observation of the training set
last_observation = tail(train_set, 1)

# Generate the forecasts by simply repeating that last row 4 times, one for each horizon (h=1, 2, 3, 4)
naive_forecasts = last_observation[rep(1, 4), ]
rownames(naive_forecasts) = c("Horizon_1", "Horizon_2", "Horizon_3", "Horizon_4")

naive_forecasts

# Alternative solution
naive_fit = train_tourism_ts%>%
  model(Naive = NAIVE(Value))
augment(naive_fit)

naive_fc = naive_fit%>%
  forecast(h = 4)
print(naive_fc, width = Inf)
```

# [Exercise 4]{style="font-size:20px;"}

If we must choose from the list provided (MAE, Average error, MAPE and
RMSE), MAPE is the only suitable measure for combining results across
these 518 series because it normalizes the errors, preventing the
large-volume tourism destinations from skewing the overall performance
assessment. Indeed, we must choose a metric that is scale-independent.
Mean Absolute Percentage Error is suitable because calculates the error
as a percentage of the actual value; this makes it scale-independent. It
only works if the data contains no zeros values and as we have checked
before in our dataset there are no zeros, so MAPE is a valid and strong
measure of forecast accuracy.

# [Exercise 5]{style="font-size:20px;"}

```{r MAPE}
# MAPE of the naive forecasts for training set (In-sample accuracy)
MAPE_train = accuracy(naive_fit)%>%
  select(Series, .type, .model, MAPE)
MAPE_train

# MAPE of the naive forecasts for validation set Out-of-sample accuracy
MAPE_test = accuracy(naive_fc, tourism_ts)%>% # Compare forecasts vs actual values in the full dataset
  select(Series, .type, .model, MAPE)
MAPE_test
```

# [Exercise 6]{style="font-size:20px;"}

```{r MASE}
# MASE of the naive forecasts for training set (In-sample accuracy)
MASE_train = accuracy(naive_fit)%>%
  select(Series, .type, .model, MASE)
MASE_train

# MASE of the naive forecasts for validation set (Out-of-sample accuracy)
MASE_test = accuracy(naive_fc, tourism_ts)%>% # Compare forecasts vs actual values in the full dataset
  select(Series, .type, .model, MASE)
MASE_test
```

The performance measure *Mean Absolute Scaled Error*(MASE) is widely
considered one of the best metrics for comparing forecast accuracy
across multiple time series. Like MAPE, MASE is independent of the scale
of the data. Unlike MAPE, which becomes infinite or undefined if an
observation is 0, MASE works perfectly fine with zero values in the time
series (as long as the series isn't completely constant). It penalizes
positive and negative errors equally and it has a very clear
benchmark:<br> - MASE \< 1: the model is better than the in-sample Naive
forecast;<br> - MASE \> 1: the model is worse than the in-sample Naive
forecast;<br> - MASE = 1: the model performs effectively the same as a
Naive forecast.

# [Exercise 7]{style="font-size:20px;"}

```{r MAPE scatterplot}
data_MAPE = left_join(MAPE_train, MAPE_test, by = 'Series')
data_MAPE

plot_MAPE = ggplot(data_MAPE, aes(x = MAPE.x, y = MAPE.y)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, col = "red", linetype = "dashed") +
  labs(title = "MAPE Scatterplot: Training vs Validation", x = "Training MAPE (%)", 
       y = "Validation MAPE (%)") +
  theme_minimal()
print(plot_MAPE)
```

<br> In the MAPE scatterplot, most points lie above the red dashed line
(bisector of the first and third quadrants) and this means the error in
the validation period (last 4 years) is generally higher than in the
training period. There seems to be a moderate positive correlation in
MAPE which means series that were "hard" to forecast in the past (high
Training MAPE) tended to remain "hard" to forecast in the future (high
Validation MAPE), though the relationship is noisy.

```{r MASE scatterplot}
data_MASE = left_join(MASE_train, MASE_test, by = 'Series')
data_MASE

plot_MASE = ggplot(data_MASE, aes(x = MASE.x, y = MASE.y)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, col = "blue", linetype = "dashed") +
  labs(title = "MASE Scatterplot: Training vs Validation", x = "Training MASE", 
       y = "Validation MASE") +
  theme_minimal()
print(plot_MASE)
```

<br> The MASE scatterplot is a vertical line at x = 1.0 because for a
Naive forecast the "Training MASE" is mathematically defined as exactly
1: while the denominator of MASE is the error of the naive method
in-sample, the numerator is also the error of the naive method
in-sample, thus the ratio is always 1. This tells us that the "Training
MASE" is not a useful metric for comparing the variability of series
when using the Naive method itself. It only serves as the baseline.
Furthermore it seems that the validation period contains more volatility
or a structural break compared to training period.

Performance ranges widely across series: there is high variance due to
the enormous variation between series. Regarding MAPE scatterplot, some
series are very stable (errors \< 5%), while others are extremely
erratic (errors \> 100%). On the other hand, in MASE scatterplot the
values range from roughly 0 (series that became much easier to forecast)
to over 40 (series that completely broke from their historical pattern
and they are harder to forecast).

# [Exercise 8]{style="font-size:20px;"}

The competition winner, Lee Baker, used an ensemble of three
methods:<br> - Naive forecasts multiplied by a constant trend
(global/local trend: ”globally tourism has grown ”at a rate of 6%
annually.”);<br> - Linear regression;<br> - Exponentially-weighted
linear regression.

## [Letter a]{style="font-size:20px;"}

Exact formula used for generating the first method is
$F_{t+k} = y_t \times (1.06)^k$ where:<br> - $F_{t+k}$ is the forecast
for k years ahead; <br> - $y_t$ is the last observed value in the
training set (the Naive component);<br> - $1.06$ represents the 6%
annual growth rate mentioned in the text (derived from the domain
knowledge that "globally tourism has grown at a rate of 6%
annually");<br> - $k$ is the forecast horizon (number of years into the
future).<br>

## [Letter b]{style="font-size:20px;"}

While standard Naive forecasting assumes the future will be identical to
the last observation ($F_{t+k} = y_t$), this "Naive with Trend" method
assumes that the last observation is the baseline, but the series will
continue to grow exponentially at a constant global rate of 6% per year.
Standard Naive forecasts assume a random walk (today is the best
predictor of tomorrow) and this works well for stock prices, which can
go up or down with equal probability. However, tourism data often
follows a Geometric Random Walk with Drift because tourism is rarely a
stationary process. Indeed, over the last several decades, it has been a
growth industry and international tourism arrivals have historically
grown as global GDP rises and the middle class expands, disposable
income for travel increases.

## [Letter c]{style="font-size:20px;"}

The model equation for this data could be
$Y_t = \beta_0 + \beta_1 t + \epsilon_t$, where:<br> - $Y_t$ is the
tourism metric (e.g., inbound tourism numbers) at time $t$;<br> - $t$ is
the time index and in a univariate setting without external data (like
GDP or prices), we use "Time" as a proxy for all the factors that cause
tourism to grow over the long term;<br> - $\beta_0$ is the intercept to
the y-axis (the theoretical baseline level of tourism at time 0);<br> -
$\beta_1$ is the marginal effect which represents the average annual
increase (or decrease) in tourism metric due to one unit increase of the
independent variable, keeping fixed all the other variables;<br> -
$\epsilon_t$ is the error term (residuals). For forecasting purposes, we
require that residuals have zero mean and are uncorrelated between them
and also with the predictors.<br>

Autoregressive model (AR) could be an alternative solution. In this
approach, we can use multiple regression with the past values of the
variable to predict the current value:
$y_t = \beta_0 + \beta_1 y_{t-1} + \beta_2 y_{t-2} + \dots + \epsilon_t$,
where predictors are the lagged values of $y_t$ and the dependent
variable is the current tourism demand. Although this method is correct,
applying it to our `tourism_data` is risky for two reasons:<br> - our
data has a strong upward trend in most of the series and AR model often
struggle with non-stationary data that grows over time;<br> - every time
we add a lag, we lose an observation from the start of our training set.
It works great for the longer series, but it will fail completely for
the shorter series because the creation of lag variables consumes the
few data points we have in the training set.

For these reasons the Simple Linear Regression is more robust for short,
trended series because it doesn't require "lagging" (and thus losing)
the early data points.

## [Letter d]{style="font-size:20px;"}

```{r linear regression}
# We filter for only the first 5 series on the long tsibble 
subset_data = tourism_ts %>%
  filter(Series %in% c("Y1", "Y2", "Y3", "Y4", "Y5"))

# Training set: all data except the last 4 years
train_subset = subset_data %>%
  filter(Time <= max(Time) - 4)

# we fit the Linear Regression Model (TSLM) on training set
fit_subset = train_subset %>%
  model(lm = TSLM(Value ~ Time))
report(fit_subset)

# We forecast the next 4 years (validation period)
fc_subset = fit_subset %>%
  forecast(h = 4)
print(fc_subset, width = Inf)

# We compare the forecasts against the actual values in the full dataset
accuracy_results = accuracy(fc_subset, subset_data) %>%
  select(Series, .model, .type, RMSE, MAE, MAPE, MASE)
accuracy_results
```

## [Letter e]{style="font-size:20px;"}

Using high-order polynomials (e.g, 5th order) to forecast time series is
dangerous because of their behavior outside the training set.
Polynomials of order $k > 1$ tend to explode towards $+\infty$ or
$-\infty$ very rapidly as you move away from the range of our data
(extrapolation). A 5th-order polynomial might wiggle to fit every bump
perfectly. However, for the next 4 years (forecast horizon), the curve
might shoot vertically upwards or crash to negative values, producing
absurd forecasts due to overfitting. This phenomenon often occurs with
too complex models or insufficient datasets, performing excellently on
training data but poorly at making predictions on novel data, leading to
poor generalization ability and making the model unreliable.

Using $R^2$ for blending the results essentially rewards overfitting. In
trended time series, $R^2$ is often artificially high (e.g., $R^2$ \>
0.90) simply because there is a trend and it doesn't punish
autocorrelation in the residuals. A model can have $R^2 = 0.99$ and
still produce biased, poor forecasts because it ignores the serial
dependence of the errors. Relying on it for blending the results of the
prediction, gives a false sense of confidence in the model.

## [Letter f]{style="font-size:20px;"}

If we were to consider exponential smoothing, the types that are
reasonable candidates:<br> - Holt's linear trend (A, A, N);<br> - Damped
trend method (A, Ad, N);<br> - previous models with multiplicative error
that are ETS(M, A, N) and ETS(M, Ad, N).

## [Letter g]{style="font-size:20px;"}

Feature-based FORecast Model Averaging (FFORMA), developed by Monash
University researchers, is a really effective approach which leads to
easier automation of the ensemble step. We start with extracting
features for every time series, calculating the main
characteristics(e.g., using feasts in R). Then, we need to use a machine
learning model like XGBoost that builds decision trees sequentially,
where each tree corrects the errors of the previous one. This focuses
heavily on reducing bias and it can directly optimize complex custom
loss functions (like the forecasting error minimization). The inputs are
the features of the series, while the outputs are the optimal weights
for the base models. The meta-learner adapts to each specific series
automatically and we have 518 custom ensembles generated by the
algorithm.<br> Also we could use another machine learning model like
Random Forest that leverages two forms of randomness to achieve low
variance and low bias; it is based on bootstrap aggregating (bagging) to
improve the accuracy of individual trees, and on random input selection
during node splitting to further enhance diversity.

## [Letter h]{style="font-size:20px;"}

In a real-life tourism forecasting scenario we may have different
weighting because high-volume/high-value series should matter more than
other series. The cost of erratic forecasting is higher, thus
under-forecasting might mean lost revenue (opportunity costs), while
over-forecasting might mean wasted budget.The competition focused on
minimizing the average MAPE of the next four values across all 518
series, while in a real-life tourism forecasting scenario MAE or RMSE
are often useful because they are scale dependent and consider the
physical number of people (e.g., seats on planes, visitor nights or
inbound tourism number to one country from another country) that matters
more than the percentage.<br> In the forecasting process the steps that
would likely be different:<br> - building a hierarchy where we could
forecast "Total Country" first, then break it down by Region, then by
City; we would ensure that the sum of the regional forecasts equals the
national forecast. This ensures consistency for decision-makers at
different levels;<br> - including explanatory variables that affect
tourism such as GDP of origin countries, exchange rates, marketing spend
or special events organized in the country;<br> - preferring a simpler
and more interpretable model (like Simple Linear Regression) because
stakeholders need to understand why the forecast is going up or down.
